---
title: "brainchart.io (Beginner Friendly)"
output: html_document
---

#  Overview
This document helps you analyze brain MRI data using R
The guide will walk you step by step through:
1. Loading the right libraries
2. Reading in brain scan and demographic data
3. Estimating brain age based on brain volume
4. Plotting the results

```{r setup, include=FALSE}
# This sets options so the code and its results are shown in the final document.
knitr::opts_chunk$set(echo = TRUE)
```## Step 1: Load Your Toolboxes (Libraries)

```{r}
library(readxl)      # For reading Excel files
library(openxlsx)    # Another way to work with Excel files
library(stringr)     # For handling text
library(dplyr)       # For organizing and summarizing data
library(readr)       # For reading CSV files
```

---

## Step 2: Load the Brain Age Template

We need a file that contains typical brain values used to estimate brain age.

```{r}
# Load a CSV file (spreadsheet with comma-separated values)
template_brainage <- read_csv("/data/raw/brainage_template.csv")

# Purpose of this line: “Save the contents of that file in a container named template_brainage”
```

---

## Step 3: Load the Subject Information

This file contains personal information like subject ID, age, sex, etc.

```{r}
# Load a spreadsheet file with subject demographic info
demo <- read_excel("/data/raw/subject_info.xlsx")

# Now we have the data stored in something called 'demo'
```

---

## Step 4: Load Brain Measurements

We load brain structure data from two files: 'aseg' and 'aparc'.

```{r}
# Load aseg data (volume of brain structures)
aseg <- read_csv("/data/raw/aseg_volumes.csv")

# Load aparc data (more brain structure measurements)
aparc <- read_csv("/data/raw/aparc_data.csv")
```

---

## Step 5: Combine the Data

#Now, let’s merge (combine) the files so we can analyze them all together.

```{r}
# Combine demographic info with aseg volumes using a common ID column
merged_data <- merge(demo, aseg, by = "ID")

# Add aparc data too
merged_data <- merge(merged_data, aparc, by = "ID")
```

---

## Step 6: Predict Brain Age

We use the brain age template to estimate how old a person’s brain “looks.”

```{r}
# This is just an example – real brain age prediction is more complex
merged_data$predicted_age <- merged_data$BrainVolume * 0.001 + 50

# This line makes a new column called 'predicted_age' using a simple formula
```

---

## Step 7: Visualize the Results

Let’s plot predicted brain age against actual age.

```{r}
plot(merged_data$Age, merged_data$predicted_age,
     main = "Predicted vs Actual Age",
     xlab = "Actual Age",
     ylab = "Predicted Brain Age",
     col = "blue", pch = 16)
```

This shows if brain age predictions match real age.

---

