---
title: "centile_brain_subvol"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Overview

This R Markdown script processes neuroimaging outputs and populates CentileBrain templates (https://centilebrain.org/#/model) based on freesurfer data.

Files and columns are standardized for RBC conventions..


```{r cars}
# Load Required Packages

library(readxl)
library(openxlsx)
library(stringr)
library(dplyr)
library(readr)
```

```{bash}
# Create main project directory (run once)
mkdir -p centilebrain_BrainAGE

# Navigate into your project folder (adjust path)
cd /path/to/your/centilebrain_BrainAGE

# Create subfolders
mkdir -p Cortical_Measures
mkdir -p output
mkdir -p templates

# Move files to correct folders (update filenames as needed)

# Move the main template, demographics, aseg files into main folder
mv ~/Downloads/<centilebrain_template_file.xlsx> ./ 
mv ~/Downloads/<demographics_file.tsv> ./
mv ~/Downloads/<aseg.tsv> ./

# Move subcortical and cortical template excels into templates folder
mv ~/Downloads/subcortical-volume-female.xlsx templates/
mv ~/Downloads/subcortical-volume-male.xlsx templates/
mv ~/Downloads/cortical-thickness-female.xlsx templates/
mv ~/Downloads/cortical-thickness-male.xlsx templates/
mv ~/Downloads/surface-area-female.xlsx templates/
mv ~/Downloads/surface-area-male.xlsx templates/

# Move thickness and surface area CSVs into Cortical_Measures folder
mv ~/Downloads/CorticalMeasuresENIGMA_ThickAvg.csv Cortical_Measures/
mv ~/Downloads/CorticalMeasuresENIGMA_SurfAvg.csv Cortical_Measures/

# Confirm structure
tree
```

## Import the necessary files

This is for all of the centile brain centile scores. 

```{r pressure, echo=FALSE}
#Load template excel file
template_subvol_fem <- read_excel("templates/subcortical-volume-female.xlsx")
template_subvol_male <- read_excel("templates/subcortical-volume-male.xlsx")
template_thick_fem <- read_excel("templates/cortical-thickness-female.xlsx")
template_thick_male <- read_excel("templates/cortical-thickness-male.xlsx")
template_area_fem <- read_excel("templates/surface-area-female.xlsx")
template_area_male <- read_excel("templates/surface-area-male.xlsx")

# Assigning variables for sex
sex_female <- "Female"

sex_male <- "Male"

#Load Aseg file
aseg_data <- read_tsv("aseg.tsv")

#Load thickness file
thickness_data <- read_csv("Cortical_Measures/CorticalMeasuresENIGMA_ThickAvg.csv")

#Load surface area file
surfacearea_data <- read_csv("Cortical_Measures/CorticalMeasuresENIGMA_SurfAvg.csv")

#Load demographics file
demographics_df <- read_tsv("demographics_file.tsv")

# Rename demographic columns to match RBC conventions
demographics_df <- demographics_df %>%
  rename(
    participant_id = subject_id,
    age = age_scan,
    sex = assigned_sex_at_birth,
    wave = time_point
  )
```

# Standardize column names

```{r}
#Rename ID columns to common names
names(aseg_data)[1] <- "participant_id"
names(thickness_data)[1] <- "participant_id"
names(surfacearea_data)[1] <- "participant_id"

```

#Function to clean demographic subject_IDs and then rename subject_id to participant_id

```{r}
#Clean demographic subject IDS
clean_subject_ids <- function(df, id_col, prefix_to_remove = NULL, remove_underscores = TRUE) {
  clean_ids <- df[[id_col]]

  # Remove custom prefix if specified
  if (!is.null(prefix_to_remove)) {
    clean_ids <- str_replace(clean_ids, paste0("^", prefix_to_remove), "sub-")
  }

  # Remove underscores if requested
  if (remove_underscores) {
    clean_ids <- str_replace_all(clean_ids, "_", "")
  }

  df$clean_subject_id <- clean_ids
  return(df)
}


# Apply the function to your data frame
demographics_df <- clean_subject_ids_demo(demographics_df, "subject_id")

# Check the results
head(demographics_df$clean_subject_id)

template_subvol_fem <- template_subvol_fem %>%
  rename(participant_id = SubjectID)
template_subvol_male <- template_subvol_male %>%
  rename(participant_id = SubjectID)
template_thick_fem <- template_thick_fem %>%
  rename(participant_id = SubjectID)
template_thick_male <- template_thick_male %>%
  rename(participant_id = SubjectID)
template_area_fem <- template_area_fem %>%
  rename(participant_id = SubjectID)
template_area_male <- template_area_male %>%
  rename(participant_id = SubjectID)

```

#Standardize Brain Region Column Names
## Need to change names of the some of the variables to match the template file. 

```{r}
# Subcortical volumes
standardize_aseg_columns <- function(col_names) {
  # Replace full names with abbreviations for hippocampus and thalamus
  col_names <- str_replace(col_names, "Thalamus-Proper", "thal")
  col_names <- str_replace(col_names, "Left-", "L")
  col_names <- str_replace(col_names, "Right-", "R")
  col_names <- str_replace_all(col_names, "-", "")
  col_names <- str_replace(col_names, "Hippocampus", "hippo")
  col_names <- str_replace(col_names, "Caudate", "caud")
  col_names <- str_replace(col_names, "Putamen", "put")
  col_names <- str_replace(col_names, "Pallidum", "pal")
  col_names <- str_replace(col_names, "Amygdala", "amyg")
  col_names <- str_replace(col_names, "Accumbensarea", "accumb")
  col_names <- str_replace(col_names, "EstimatedTotalIntraCranialVol", "ICV")
    
  
  return(col_names)
}

colnames(aseg_data) <- standardize_aseg_columns(colnames(aseg_data))

# Cortical thickness and surface area
standardize_col_names <- function(col_names) {
  # Replace full names with abbreviations for hippocampus and thalamus
  col_names <- str_replace(col_names, "right", "R")
  col_names <- str_replace(col_names, "left", "L")
  col_names <- str_replace_all(col_names, "entorhinal", "entorhil")
  col_names <- str_replace_all(col_names, "supramarginal", "supramargil")
  
  return(col_names)
}

colnames(thickness_data) <- standardize_col_names(colnames(thickness_data))
colnames(surfacearea_data) <- standardize_col_names(colnames(surfacearea_data))


```


1. This First filters for time point because we only want baseline scans (also prevents duplicate subject_IDs in the final output)

2. Then filters for sex and then creates a dataframe with subject Id, age, and sex. 

```{r}
#Split demographics by Sex and Filter

female_data <- demographics_df %>%
  filter(wave == "01") %>%  # First filter for the desired time_point
  filter(sex == sex_female) %>%  # Then filter for sex
  select(participant_id, age, sex)  # Keep subject ID, age, and sex for merging

male_data <- demographics_df %>%
  filter(wave == "01") %>%  # First filter for the desired time_point
  filter(sex == sex_male) %>%  # Then filter for sex
  select(clean_subject_id, age, sex)  # Keep subject ID, age, and sex for merging

```

Filter the aseg data for sex

```{r}
# Filter raw data by clean IDs
female_aseg_data_clean <- aseg_data %>%
  filter(clean_subject_id %in% female_data$clean_subject_id)

female_surfacearea_data_clean <- surfacearea_data %>%
  filter(clean_subject_id %in% female_data$clean_subject_id)

female_thickness_data_clean <- thickness_data %>%
  filter(clean_subject_id %in% female_data$clean_subject_id)

male_aseg_data_clean <- aseg_data %>%
  filter(clean_subject_id %in% male_data$clean_subject_id)

male_surfacearea_data_clean <- surfacearea_data %>%
  filter(clean_subject_id %in% male_data$clean_subject_id)

male_thickness_data_clean <- thickness_data %>%
  filter(clean_subject_id %in% male_data$clean_subject_id)

```

# Add the sex specific subject_ID, age, and sex to the template_df
## This also ensures that there are enough rows to merge the data. 

```{r}
#Fill template with subject info

update_template <- function(template_df, gendered_data) {
  # Ensure input types are correct
  template_df$clean_subject_id <- as.character(template_df$clean_subject_id)
  template_df$sex <- as.character(template_df$sex)
  template_df$age <- as.character(template_df$age)

  # Check if template_df is empty
  if (nrow(template_df) == 0) {
    # Create an empty data frame with the same structure as template_df
    template_df <- template_df[0, ]  # Retain column structure but no rows
    
    # Create a new data frame from gendered_data with required columns
    new_data <- data.frame(
      clean_subject_id = gendered_data$clean_subject_id,
      age = gendered_data$age,
      sex = gendered_data$sex,
      stringsAsFactors = FALSE
    )
    
    # Ensure consistent data types
    new_data$clean_subject_id <- as.character(new_data$participant_id)
    new_data$sex <- as.character(new_data$sex)
    new_data$age <- as.character(new_data$age)
    
    # Combine the new data into the empty template
    template_df <- bind_rows(template_df, new_data)
  } else {
    # If template_df has rows, update columns directly
    template_df$clean_subject_id <- gendered_data$participant_id
    template_df$age <- gendered_data$age
  }
  
  # Return the updated template
  return(template_df)
}

# Example application for one template
template_subvol_fem <- update_template(template_df = template_subvol_fem, gendered_data = female_data)
template_thick_fem <- update_template(template_df = template_thick_fem, gendered_data = female_data)
template_area_fem <- update_template(template_df = template_area_fem, gendered_data = female_data)

template_subvol_male <- update_template(template_df = template_subvol_male, gendered_data = male_data)
template_thick_male <- update_template(template_df = template_thick_male, gendered_data = male_data)
template_area_male <- update_template(template_df = template_area_male, gendered_data = male_data)

```

Add the default values for the SITE (1 because they were all taken at the same site on the same scanner), ScannerType, and FreeSurfer_Version. Change values if needed. 

```{r}

# Create a named list of default values for columns with missing data
default_values <- list(
  SITE = "SITE_01",
  Vendor = "Scanner Vendor",
  FreeSurfer_Version = "x.y"
)

# Update existing columns with default values if they are NA
for (col in names(default_values)) {
  if (col %in% names(template_subvol_fem)) {
    template_subvol_fem[[col]] <- ifelse(is.na(template_subvol_fem[[col]]), default_values[[col]], template_subvol_fem[[col]])
  }
}

# Update existing columns with default values if they are NA
for (col in names(default_values)) {
  if (col %in% names(template_subvol_male)) {
    template_subvol_male[[col]] <- ifelse(is.na(template_subvol_male[[col]]), default_values[[col]], template_subvol_male[[col]])
  }
}

# Update existing columns with default values if they are NA
for (col in names(default_values)) {
  if (col %in% names(template_thick_fem)) {
    template_thick_fem[[col]] <- ifelse(is.na(template_thick_fem[[col]]), default_values[[col]], template_thick_fem[[col]])
  }
}

# Update existing columns with default values if they are NA
for (col in names(default_values)) {
  if (col %in% names(template_thick_male)) {
    template_thick_male[[col]] <- ifelse(is.na(template_thick_male[[col]]), default_values[[col]], template_thick_male[[col]])
  }
}

# Update existing columns with default values if they are NA
for (col in names(default_values)) {
  if (col %in% names(template_area_fem)) {
    template_area_fem[[col]] <- ifelse(is.na(template_area_fem[[col]]), default_values[[col]], template_area_fem[[col]])
  }
}

# Update existing columns with default values if they are NA
for (col in names(default_values)) {
  if (col %in% names(template_area_male)) {
   template_area_male[[col]] <- ifelse(is.na(template_area_male[[col]]), default_values[[col]], template_area_male[[col]])
  }
}

```

Convert all the columns from the files that need to be merged to characters so that they can be merged.

```{r}

# Convert all columns in the template_df to characters
template_subvol_fem <- template_subvol_fem %>%
  mutate(across(everything(), as.character))

template_subvol_male <- template_subvol_male %>%
  mutate(across(everything(), as.character))

template_thick_fem <- template_thick_fem %>%
  mutate(across(everything(), as.character))

template_thick_male <- template_thick_male %>%
  mutate(across(everything(), as.character))

template_area_fem <- template_area_fem %>%
  mutate(across(everything(), as.character))

template_area_male <- template_area_male %>%
  mutate(across(everything(), as.character))

male_aseg_data_clean <- male_aseg_data_clean %>%
  mutate(across(everything(), as.character))

male_thickness_data_clean <- male_thickness_data_clean %>%
  mutate(across(everything(), as.character))

male_surfacearea_data_clean <- male_surfacearea_data_clean %>%
  mutate(across(everything(), as.character))

female_aseg_data_clean <- female_aseg_data_clean %>%
  mutate(across(everything(), as.character))

female_thickness_data_clean <- female_thickness_data_clean %>%
  mutate(across(everything(), as.character))

female_surfacearea_data_clean <- female_surfacearea_data_clean %>%
  mutate(across(everything(), as.character))

```

Populate the template df by joining the aseg, surface area, and thickness data. 

```{r}
process_template <- function(template_df, primary_data_clean) {
  # Perform the left join
  final_template <- left_join(template_df[, c(1, 2, 3, 4, 5, 6)], 
                                primary_data_clean, by = "clean_subject_id")
  
  # Select only the columns from the template
  final_template <- final_template %>%
    select(names(template_df))
  
 # Replace "sub-EXCLUDE" with any known QA/test subject IDs to exclude
  final_template <- final_template %>%
    filter(clean_subject_id != "sub-exclude")
  
  # Return the processed data frame
  return(final_template)
}

final_subvol_fem <- process_template(template_subvol_fem, female_aseg_data_clean)
final_subvol_male <- process_template(template_subvol_male, male_aseg_data_clean)
# For thickness data, also merge with surface area data
final_thick_fem <- process_template(template_thick_fem, 
                                    female_thickness_data_clean)
final_thick_male <- process_template(template_thick_male, 
                                     male_thickness_data_clean)
```

```{r}
female_surfacearea_data_clean <- female_surfacearea_data_clean %>%
  select(-LSurfArea, -RSurfArea)

male_surfacearea_data_clean <- male_surfacearea_data_clean %>%
  select(-LSurfArea, -RSurfArea)

final_area_fem <- left_join(template_area_fem[, 1:6], female_surfacearea_data_clean, by = "clean_subject_id")

# Select only the columns from template_df
final_area_fem <- final_area_fem %>%
  select(names(template_area_fem))  # Keeps only the columns in template_df 

final_area_male <- left_join(template_area_male[, 1:6], male_surfacearea_data_clean, by = "clean_subject_id")


# Select only the columns from template_df
final_area_male <- final_area_male %>%
  select(names(template_area_male))  # Keeps only the columns in template_df 

```


```{r}
final_subvol_fem <- na.omit(final_subvol_fem)
final_subvol_male <- na.omit(final_subvol_male)
final_thick_fem <- na.omit(final_thick_fem)
final_thick_male <- na.omit(final_thick_male)
final_area_fem <- na.omit(final_area_fem)
final_area_male <- na.omit(final_area_male)

```


```{r}
# Convert all other rows and columns to numeric
numeric_subvol_fem <- final_subvol_fem %>%
  mutate_at(vars(-c(3, 6)), ~ as.numeric(as.character(.)))

numeric_subvol_fem <- numeric_subvol_fem %>%
  rename(SubjectID = clean_subject_id)

numeric_thick_fem <- final_thick_fem %>%
  mutate_at(vars(-c(3, 6)), ~ as.numeric(as.character(.)))

numeric_thick_fem <- numeric_thick_fem %>%
  rename(SubjectID = clean_subject_id)

numeric_area_fem <- final_area_fem %>%
  mutate_at(vars(-c(3, 6)), ~ as.numeric(as.character(.)))

numeric_area_fem <- numeric_area_fem %>%
  rename(SubjectID = clean_subject_id)

numeric_subvol_male <- final_subvol_male %>%
  mutate_at(vars(-c(3, 6)), ~ as.numeric(as.character(.)))

numeric_subvol_male <- numeric_subvol_male %>%
  rename(SubjectID = clean_subject_id)

numeric_thick_male <- final_thick_male %>%
  mutate_at(vars(-c(3, 6)), ~ as.numeric(as.character(.)))

numeric_thick_male <- numeric_thick_male %>%
  rename(SubjectID = clean_subject_id)

numeric_area_male <- final_area_male %>%
  mutate_at(vars(-c(3, 6)), ~ as.numeric(as.character(.)))

numeric_area_male <- numeric_area_male %>%
  rename(SubjectID = clean_subject_id)

```
```{r}
numeric_subvol_fem <- na.omit(numeric_subvol_fem)
numeric_subvol_male <- na.omit(numeric_subvol_male)
numeric_thick_fem <- na.omit(numeric_thick_fem)
numeric_thick_male <- na.omit(numeric_thick_male)
numeric_area_fem <- na.omit(numeric_area_fem)
numeric_area_male <- na.omit(numeric_area_male)

```

```{r}
# Save populated templates to the output folder

output_subvol_fem <- "output/populated_subvol_fem.xlsx"
write.xlsx(numeric_subvol_fem, output_subvol_fem)

output_subvol_male <- "output/populated_subvol_male.xlsx"
write.xlsx(numeric_subvol_male, output_subvol_male)

output_thick_fem <- "output/populated_thick_fem.xlsx"
write.xlsx(numeric_thick_fem, output_thick_fem)

output_thick_male <- "output/populated_thick_male.xlsx"
write.xlsx(numeric_thick_male, output_thick_male)

output_area_fem <- "output/populated_area_fem.xlsx"
write.xlsx(numeric_area_fem, output_area_fem)

output_area_male <- "output/populated_area_male.xlsx"
write.xlsx(numeric_area_male, output_area_male)

# Optional confirmation messages
print(paste("Saved:", output_subvol_fem))
print(paste("Saved:", output_subvol_male))
print(paste("Saved:", output_thick_fem))
print(paste("Saved:", output_thick_male))
print(paste("Saved:", output_area_fem))
print(paste("Saved:", output_area_male))
```

# References

Bethlehem, R. A. I., Seidlitz, J., White, S. R., et al. (2022). Brain charts for the human lifespan. Nature, 604(7906), 525–533.
https://doi.org/10.1038/s41586-022-04554-y