---
title: "Developmental_brain_age"
output: html_notebook
date: 2024-10-17
---
# Overview
This is a script to create the Developmental BrainAge template (https://centilebrain.org/#/brainAGE) using freesurfer outputs.
**Please follow each step and modify the paths and options as needed.**

---

# Load the necessary libraries

```{r}
library(readxl)
library(openxlsx)
library(stringr)
library(dplyr)
library(readr)
```
```{bash}
# Create a directory to store files for BrainAGE calculation. 

# Create the main project directory Developmental_BrainAGE folder
mkdir -p Developmental_BrainAGE
cd ~/Developmental_BrainAGE

# Make folders within Developmental_BrainAGE_BrainAGE folder
mkdir -p Developmental_BrainAGE/Freesurfer_Outputs
mkdir -p Developmental_BrainAGE/output

# Confirm the structure
tree Developmental_BrainAGE
```

#Download the BrainAGE template, project demographic data, and freesurfer outputs


```{bash}
# Move template and demographic files into main project folder
mv ~/Downloads/template_brainage.xlsx Developmental_BrainAGE/
mv /path/to/your/demographic_file.tsv Developmental_BrainAGE/ #Replace /path/to/your/demographic_file.tsv with your actual file path

# Move freesurfer files into subfolder
mv ~/Downloads/lh.Schaefer2018_400Parcels.*.tsv Developmental_BrainAGE/Freesurfer_Outputs/
mv ~/Downloads/rh.Schaefer2018_400Parcels.*.tsv Developmental_BrainAGE/Freesurfer_Outputs/
```
# Load Files

# Filter for Sex ('Male' or 'Female')
sex <- "Female" # Change sex to "Male" or "Female" depending on analysis.

```{r}
# Set working directory to main project folder before running this script
setwd("~/Developmental_BrainAGE")

# Load Developmental BrainAGE template and demographic data
template_brainage <- read_excel("template_brainage.xlsx")
demographic_df <- read_tsv("study-HBN_desc-participants.tsv")

# Load parcellation files (update path if needed)
schaefer_wide_data_path <- file.path("Freesurfer_Outputs", "Schaefer2018_400Parcels_7Networks_order_wide.csv")
schaefer_400_7N_wide <- read_csv(schaefer_wide_data_path, show_col_types = FALSE)
brainmeasures <- read_csv(combined_brainmeasures.csv)
```

# Clean column names to match the Centile BrainAGE template.
# 1. Saves original column names for reference
# 2. Removes underscores from column names
# 3. Removes first n characters. The length of the columns are checked because some column names have fewer characters and thus if they are removed, it will result in empty column headers. 

```{r}
# Save the original column names for comparison with the template
original_template_column_names <- colnames(template_brainage)

# Function to clean columns
clean_columns <- function(df, remove_char_count) {
  if (is.data.frame(df) && ncol(df) > 0) {
    colnames(df) <- str_replace_all(colnames(df), "_", "")  # Remove underscores

    # Check length of each column name and conditionally remove characters
    colnames(df) <- ifelse(nchar(colnames(df)) > remove_char_count,
                           str_sub(colnames(df), remove_char_count + 1), 
                           colnames(df))  # Remove first n characters if long enough
  }
  return(df)
}
   
```

# Apply cleaning function
```{r}
lh_thickness_df <- clean_columns(lh_thick, 11)  
lh_area_df <- clean_columns(lh_surf, 11)  
rh_thickness_df <- clean_columns(rh_thick, 11)  
rh_area_df <- clean_columns(rh_surf, 11)  

# Different number of characters for the template

template_df <- clean_columns(template_brainage, 10)
```

# Rename columns in the thickness data from "thickness" to "thick" to match the column names in the template

```{r}
colnames(lh_thickness_df) <- str_replace_all(colnames(lh_thickness_df), "ThickAvg", "thick") #
colnames(rh_thickness_df) <- str_replace_all(colnames(rh_thickness_df), "ThickAvg", "thick") 

```

# Function to clean and standardize subject IDs for the thickness and surface area data

```{r}
clean_subject_ids_tsv <- function(df) {
  colnames(df)[1] <- "clean_subject_id"  # Rename the first column to 'clean_subject_id'
  df %>%
    mutate(clean_subject_id = str_extract(clean_subject_id, "[0-9]{8}$"))  # Extract the last 8 digits of subject IDs
}
```

# Apply the subject_ID cleaning function to thickness and surface area files

```{r}
lh_thick_clean <- clean_subject_ids_tsv(lh_thickness_df)
lh_surf_clean <- clean_subject_ids_tsv(lh_area_df)
rh_thick_clean <- clean_subject_ids_tsv(rh_thickness_df)
rh_surf_clean <- clean_subject_ids_tsv(rh_area_df)
```

# Define a function to clean demographic subject_ids

```{r}
clean_subject_ids_demo <- function(df, subject_id) {
  df %>%
    mutate(clean_subject_id = str_extract(!!sym(subject_id), "[0-9]{8}$"))
}
```

# Apply the cleaning function to the demographic subject_ids

# Check if subject ID column exists in template, otherwise add temp_id
if (!("clean_subject_id" %in% colnames(template_brainage))) {
  message("No subject ID column in template_brainage. Adding temporary IDs as temp_id.")
  template_brainage$temp_id <- seq_len(nrow(template_brainage))
  id_col <- "temp_id"
} else {
  id_col <- "clean_subject_id"
}
# Rename subject_id to clean_subject_id in the demographic file (to standardize the column name)

```{r}
template_df <- template_df 
demographic_df <- clean_subject_ids_demo(demographic_df, "subject_id")
```
# Rename demographic column for merging to standard clean_subject_id
# Only do this if your template has clean_subject_id, else use temp_id
if (id_col == "clean_subject_id") {
  template_df <- template_df %>% rename(clean_subject_id = subjectid)
}

# Filter the dataset to keep only baseline scans (time_point == "01") to avoid duplicate subject_IDs in the final output. 
#Then filter for sex and create a dataframe with just subject Id, age, and sex. 
```{r}
sex <- "F"  # or "M"
# Check if id_col (from template_brainage) exists in demographic_df
if (id_col %in% colnames(demographic_df)) {
  # Then rename that column to id_col (e.g. clean_subject_id to temp_id)
  gendered_data <- demographic_df %>%
    filter(wave == "01") %>%  
    filter(sex == !!sex) %>%  
    select(clean_subject_id, age, sex) %>% 
    rename(AGE = age, !!id_col := clean_subject_id)
} else {
  # Just rename age to AGE, don't try to rename ID column
  gendered_data <- demographic_df %>%
    filter(wave == "01") %>%  
    filter(sex == !!sex) %>%  
    select(clean_subject_id, age, sex) %>% 
    rename(AGE = age)
}
```

# Filter thickness and area data for female subject IDs
# Consider creating a function to filter subject_IDs

```{r}
gendered_lh_thickness_data <- lh_thick_clean %>%
  filter(clean_subject_id %in% gendered_data$clean_subject_id)

gendered_lh_area_data <- lh_surf_clean %>%
  filter(clean_subject_id %in% gendered_data$clean_subject_id)

gendered_rh_thickness_data <- rh_thick_clean %>%
  filter(clean_subject_id %in% gendered_data$clean_subject_id)

gendered_rh_area_data <- rh_surf_clean %>%
  filter(clean_subject_id %in% gendered_data$clean_subject_id)
```

# Ensure that all columns remain intact while adding the female data

```{r}

#Make sure the the subject_ids are character data types
template_df[[id_col]] <- as.character(template_df[[id_col]])


# First, check if template_df has rows
if (nrow(template_df) == 0) {
  # Create an empty data frame with the same structure as template_df but add the subject IDs and age
  # Keep all the original columns, but add the new data
  template_df <- template_df[0, ]  # Retain column structure but no rows
  
  # Now create a new data frame that only updates clean_subject_id and AGE
 new_data <- data.frame(clean_subject_id = gendered_data[[id_col]],
                       AGE = gendered_data$AGE,
                       stringsAsFactors = FALSE)
  
  # Bind the new data to the empty template, ensuring columns remain intact
  template_df <- bind_rows(template_df, new_data)
} else {
  # If template_df already has rows, just update clean_subject_id and AGE columns
  template_df[[id_col]] <- gendered_data[[id_col]]
  template_df$AGE <- gendered_data$AGE
}

```


# Check if subject IDs from both dataframes match

```{r}
common_ids <- intersect(template_df[[id_col]], gendered_lh_thickness_data$clean_subject_id)
print(paste("Number of matching subject IDs:", length(common_ids)))
```

# Make all of the logical variables into characters

```{r}

# Convert all columns in the template_df to characters
template_df <- template_df %>%
  mutate(across(everything(), as.character))

gendered_data <- gendered_data %>%
  mutate(across(everything(), as.character))

gendered_lh_thickness_data <- gendered_lh_thickness_data %>%
  mutate(across(everything(), as.character))

gendered_lh_area_data <- gendered_lh_area_data %>%
  mutate(across(everything(), as.character))

gendered_rh_thickness_data <- gendered_rh_thickness_data %>%
  mutate(across(everything(), as.character))

gendered_rh_area_data <- gendered_rh_area_data %>%
  mutate(across(everything(), as.character))
```

# Populate the template with female subject data and only keep the column names that are in the template. 
#Remove any rows with NA, the calculator will not work if there are any NA values. Then revert back to the original column names. 
#Ensure that all the relevant data is properly merged.

```{r}
final_template <- template_df %>% 
  select(clean_subject_id, AGE) %>% 
  left_join(gendered_lh_thickness_data, by = "clean_subject_id") %>%
  left_join(gendered_lh_area_data, by = "clean_subject_id") %>%
  left_join(gendered_rh_thickness_data, by = "clean_subject_id") %>%
  left_join(gendered_rh_area_data, by = "clean_subject_id")

# Select only the columns from template_df
final_template <- final_template %>%
  select(names(template_df))  # Keeps only the columns in templa5te_df  

# Remove columns with NA
final_template <- final_template %>%
  filter(!is.na(.[[3]]))

# Revert back to original column names found in the template
colnames(final_template)[1:length(original_template_column_names)] <- original_template_column_names
```

# Output a .csv file that contains the subject ID and the age so that they can be used for the analysis after

```{r}
output_csv_file <- paste0("../data/processed/", tolower(sex), "_subID_developmental_template.csv")
write.csv(final_template, file = output_csv_file, row.names = FALSE)
```


# Change all the data types to numeric and remove the first column of the subject ids because that was not in the original template. 
#Make sure all the data types are numeric because the calculator will not work if they are different data types. 

```{r}
# Extract the first column
first_column <- final_template[, 1]

# Convert all other rows and columns to numeric, starting from row 2 and column 2
numeric_data <- final_template[, -1] %>%
  mutate(across(everything(), ~ suppressWarnings(as.numeric(as.character(.)))))

```

# Save the final populated template into a new Excel file

```{r}

output_file <- paste0("../data/processed/", tolower(sex), "_populated_developmental_template.xlsx")
write.xlsx(numeric_data, output_file)

print(paste("Template has been populated and saved as", output_file))
```

