---
title: "Developmental_brain_age"
output: html_notebook
date: 2024-10-17
---
# Overview
This is a script to create the Developmental BrainAge template (https://centilebrain.org/#/brainAGE) using freesurfer outputs.
**Please follow each step and modify the paths and options as needed.**

---

# Load the necessary libraries

```{r}
library(readxl)
library(openxlsx)
library(stringr)
library(dplyr)
library(readr)
```

```{bash}
# Create a directory to store files for BrainAGE calculation. 

# Create the main project directory Developmental_BrainAGE folder
mkdir -p Developmental_BrainAGE
cd ~/Developmental_BrainAGE

# Make folders within Developmental_BrainAGE_BrainAGE folder
mkdir -p Developmental_BrainAGE/output

# Confirm the structure
tree Developmental_BrainAGE
```

#Download the BrainAGE template, project demographic data, and freesurfer outputs


```{bash}
# Move template and demographic files into main project folder
mv ~/path/to/your/template_brainage.xlsx Developmental_BrainAGE/
mv ~/path/to/your/demographic_file.tsv Developmental_BrainAGE/ #Replace with your actual file path

# Move parcellation data and combined brain measures data into directory
mv ~path/to/your/Schaefer2018_400Parcels.*.csv Developmental_BrainAGE
mv ~path/to/your/combined_brainmeasures.*.tsv Developmental_BrainAGE
```
# Load Files

# Filter for Sex ('Male' or 'Female')
sex <- "Female" # Change sex to "Male" or "Female" depending on analysis.

```{r}
# Set working directory to main project folder before running this script
setwd("~/Developmental_BrainAGE")

# Load Developmental BrainAGE template and demographic data
template_brainage <- read_excel("template_brainage.xlsx")
demographic_df <- read_tsv("study-HBN_desc-participants.tsv")

# Load parcellation files (update path if needed)
atlas_data <- read_csv("Schaefer2018_400Parcels_7Networks_order_wide.csv") 
```

# Clean column names to match the Developmental BrainAGE template.
# 1. Saves original column names for reference
# 2. Match column names in the dataset to match template column names
```{r}
# Save the original column names for comparison with the template
original_template_column_names <- colnames(template_brainage)

# Function to clean columns
# Fix data column names to match template column names

atlas_renamed <- atlas_data %>%
  rename_with(~ {
    name <- .x # Get the current column name

    # Apply the first replacement only if the pattern matches
    name_transformed <- if_else(
      str_detect(name, "^(lh_|rh_)(7Networks_)(L[HR]_).*_(ThickAvg|SurfArea)$"),
      str_replace(name, "^(lh_|rh_)(7Networks_)(L[HR]_)", "X\\2\\3"),
      name # If pattern doesn't match, keep the original name
    )

    # Apply the ThickAvg to thick replacement
    name_transformed <- str_replace(name_transformed, "_ThickAvg$", "thick")

    # Apply the SurfArea to area replacement
    name_transformed <- str_replace(name_transformed, "_SurfArea$", "area")

    return(name_transformed) # Return the potentially transformed name
  },
  # Apply this renaming only to columns that start with lh_7Networks_ or rh_7Networks_and end with ThickAvg or SurfArea, as specified by your needs.
  .cols = c(
    matches("^(lh_|rh_)7Networks_.*_ThickAvg$") |
    matches("^(lh_|rh_)7Networks_.*_SurfArea$")
  )
)

```
# Function to clean and standardize subject IDs for the thickness and surface area data

```{r}
atlas_clean <- atlas_renamed %>%
  mutate(clean_subject_id = str_remove(subject_id, "^sub-"))
```

# Prepare demographic data
```{r}
demographic_clean <- demographic_df %>%
  mutate(clean_subject_id = participant_id)
```
# Prepare demographic data filtered for baseline (wave == "1") and sex (e.g., "Female"), selecting clean_subject_id and age
```{r}
gendered_data <- demographic_clean %>%
  filter(wave == "1", sex == "Female") %>%
  select(clean_subject_id, age) %>%
  rename(AGE = age)
```

# Create or update the template data frame with subject IDs and age from demographic data
```{r}
if (!("clean_subject_id" %in% colnames(template_brainage))) {
  message("Template has no subject IDs — creating new one.")
  
  # Initialize empty template with same structure
  template_df <- data.frame(matrix(ncol = ncol(template_brainage), nrow = nrow(gendered_data)))
  colnames(template_df) <- colnames(template_brainage)
  
  # Insert subject IDs and ages
  template_df$clean_subject_id <- gendered_data$clean_subject_id
  template_df$AGE <- gendered_data$AGE
} else {
  template_df <- template_brainage
}
```

# Filter atlas data for subjects present in your demographic subset (e.g., females, baseline)
```{r}
gendered_atlas_data <- atlas_clean %>%
  filter(clean_subject_id %in% gendered_data$clean_subject_id)
```


# Ensure that all columns remain intact while adding the female data

```{r}
# Define ID column name
id_col <- "clean_subject_id"

# Ensure subject IDs are character type
template_df[[id_col]] <- as.character(template_df[[id_col]])
gendered_data[[id_col]] <- as.character(gendered_data[[id_col]])

# If template_df is empty (0 rows), initialize with gendered_data and proper columns
if (nrow(template_df) == 0) {
  # Create an empty data frame with template_df columns but no rows
  template_df <- template_df[0, ]

  # Create new data frame with subject IDs and AGE
  new_data <- gendered_data %>%
    select(all_of(c(id_col, "AGE")))

  # Bind new_data to template_df (other columns will be NA)
  template_df <- bind_rows(template_df, new_data)
} else {
  # If template_df already has rows, merge AGE from gendered_data by subject ID
  template_df <- template_df %>%
    select(-AGE) %>%                     # Remove existing AGE to avoid duplicates
    left_join(gendered_data %>% select(all_of(c(id_col, "AGE"))), by = id_col)
}
```

# Ensure all data are characters to avoid merge issues
# Convert everything to character
```{r}
template_df <- template_df %>% mutate(across(everything(), as.character))
gendered_data <- gendered_data %>% mutate(across(everything(), as.character))
gendered_atlas_data <- gendered_atlas_data %>% mutate(across(everything(), as.character))
```

# Check if subject IDs from both dataframes match
```{r}
common_ids <- intersect(template_df$clean_subject_id, gendered_data$clean_subject_id)
print(paste("Number of matching subject IDs:", length(common_ids)))
```

# Merge Subject ID and Age into template AND brain measures
```{r}
# Start `final_template` with subject IDs and their ages
final_template <- gendered_data %>%
  select(clean_subject_id, AGE)

# Get the brain measure columns from your `template_brainage`
brain_measure_cols_from_template <- setdiff(colnames(template_brainage), c("clean_subject_id", "AGE"))

# Select relevant brain measure columns from `gendered_atlas_data` for merging
atlas_data_for_merge <- gendered_atlas_data %>%
  select(clean_subject_id, all_of(brain_measure_cols_from_template))

# Merge the brain measure data into `final_template`
final_template <- final_template %>%
  left_join(atlas_data_for_merge, by = "clean_subject_id")

# Add any other columns from the original template that aren't yet in `final_template`
all_template_cols <- colnames(template_brainage)
missing_in_final <- setdiff(all_template_cols, colnames(final_template))

if (length(missing_in_final) > 0) {
  final_template[missing_in_final] <- NA
}

# Reorder final_template columns to match the original template_brainage order
final_template <- final_template[, all_template_cols]
```

# Output a .csv file that contains the subject ID and the age so that they can be used for the analysis after

```{r}
sex <- "female"  # Define sex variable

output_csv_file <- paste0("../data/processed/", tolower(sex), "_subID_developmental_template.csv")
write.csv(final_template, file = output_csv_file, row.names = FALSE)

```

# Change all the data types to numeric and remove the first column of the subject ids because that was not in the original template. 
```{r}
# Store first column (subject IDs)
first_column <- final_template[, 1]

# Remove ID column and convert rest to numeric
numeric_data <- final_template %>%
  select(-all_of(id_col)) %>%
  mutate(across(everything(), ~ suppressWarnings(as.numeric(.))))
```

# Save the final populated template into a new Excel file
```
```{r}
# Define output directory and ensure path is expanded
output_dir <- path.expand("~/Developmental_BrainAGE/output/")

# Create output file path
output_file <- paste0(output_dir, tolower(sex), "_populated_developmental_template.xlsx")

# Write the Excel file
write.xlsx(numeric_data, output_file)

# Print confirmation with full path
print(paste("Template has been populated and saved as", output_file))
```

