---
title: "Developmental_brain_age"
output: html_notebook
date: 2024-10-17
---
# Overview
This is a script to create the Developmental BrainAge template (https://centilebrain.org/#/brainAGE) using freesurfer outputs.
**Please follow each step and modify the paths and options as needed.**

---

# Load the necessary libraries

```{r}
library(readxl)
library(openxlsx)
library(stringr)
library(dplyr)
library(readr)
library(rlang) 
```

```{bash}
# Create a directory to store files for BrainAGE calculation. 

# Create the main project directory Developmental_BrainAGE folder
mkdir -p Developmental_BrainAGE
cd ~/Developmental_BrainAGE

# Make folders within Developmental_BrainAGE_BrainAGE folder
mkdir -p Developmental_BrainAGE/output

# Confirm the structure
tree Developmental_BrainAGE
```

#Download the BrainAGE template, project demographic data, and freesurfer outputs


```{bash}
# Move template and demographic files into main project folder
mv ~/path/to/your/template_brainage.xlsx Developmental_BrainAGE/
mv ~/path/to/your/demographic_file.tsv Developmental_BrainAGE/ #Replace with your actual file path

# Move parcellation data and combined brain measures data into directory
mv ~path/to/your/Schaefer2018_400Parcels.*.csv Developmental_BrainAGE
mv ~path/to/your/combined_brainmeasures.*.tsv Developmental_BrainAGE
```
# Load Files

# Filter for Sex ('Male' or 'Female')
sex <- "Female" # Change sex to "Male" or "Female" depending on analysis.

```{r}
# Set working directory to main project folder before running this script
setwd("~/Developmental_BrainAGE")

# Load Developmental BrainAGE template and demographic data
template_brainage <- read_excel("template_brainage.xlsx")
demographic_df <- read_tsv("study-HBN_desc-participants.tsv")

# Load parcellation files (update path if needed)
atlas_data <- read_csv("Schaefer2018_400Parcels_7Networks_order_wide.csv") 

# Define sex for filtering later
sex_filter <- "Female" # Change sex_filter to "Male" or "Female" depending on analysis.
```

# Clean column names to match the Developmental BrainAGE template and prepare Atlas data.
```{r}
# Define the renaming function
# This function is designed to take a SINGLE column name and return its new name.
convert_atlas_col_name <- function(name) {
  # This regex extracts 5 key parts from Schaefer 7Networks column names:
  # 1. (lh|rh): Initial hemisphere (e.g., 'lh_') - for pattern matching.
  # 2. (LH|RH): Main hemisphere (e.g., 'LH_') - used directly in the new name.
  # 3. ([A-Za-z_]+): Network/parcel name (e.g., 'Cont_Cing') - keeps original capitalization.
  # 4. ([0-9]+): Parcel number (e.g., '1').
  # 5. (.*): Measurement type (e.g., 'SurfArea', 'CurvInd').

  match_schaefer <- str_match(name, "^(lh|rh)_7Networks_(LH|RH)_([A-Za-z_]+)_([0-9]+)_(.*)$")

  # Check if the name matches the expected Schaefer atlas pattern
  if (!is.na(match_schaefer[1,1])) {
    # Extract components from the regex match
    internal_hemi <- match_schaefer[1, 3]
    network_parcel_name <- match_schaefer[1, 4]
    parcel_num <- match_schaefer[1, 5]
    measure_type <- tolower(match_schaefer[1, 6])

    # Build the core of the new name: X7Networks_HEMI_NetworkParcel_Number
    new_name_base <- paste("X7Networks", internal_hemi, network_parcel_name, parcel_num, sep = "_")

    # Append 'area' or 'thick' for specific measures, otherwise append original measure type
    if (measure_type == "surfarea") {
      new_name_final <- paste0(new_name_base, "area")
    } else if (measure_type == "thickavg") {
      new_name_final <- paste0(new_name_base, "thick")
    } else {
      new_name_final <- paste0(new_name_base, tolower(measure_type))
    }
    return(new_name_final)

  } else {
    # If no match, return the original name unchanged (e.g., 'session_id').
    return(name)
  }
}

# Identify the columns that need renaming
cols_to_rename <- names(atlas_data)[
  startsWith(names(atlas_data), "lh_") | startsWith(names(atlas_data), "rh_") &
  grepl("7Networks_", names(atlas_data), ignore.case = TRUE)
]

# Apply the renaming function to these column names
new_col_names <- sapply(cols_to_rename, convert_atlas_col_name, USE.NAMES = FALSE)

# Create a named vector for renaming: c(new_name = old_name, ...)
renaming_vector <- setNames(cols_to_rename, new_col_names)

# Apply the renaming and create clean_subject_id 
atlas_processed <- atlas_data %>%
  rename(!!!renaming_vector) %>% # Apply the Schaefer atlas column renaming
  mutate(clean_subject_id = str_remove(session_id, "^sub-")) %>% # Create clean_subject_id from session_id
  select(-session_id) # Remove the original session_id column
```
# Prepare Demographic Data

```{r}
demographic_processed <- demographic_df %>%
  mutate(clean_subject_id = participant_id) %>% # Ensure clean_subject_id matches atlas data format
  filter(wave == "1", sex == sex_filter) %>% # Filter by wave to avoid duplicates and the specified sex_filter
  select(clean_subject_id, age) %>% # Select only necessary columns
  rename(AGE = age) %>% # Rename 'age' to 'AGE' as per template
  # Ensure clean_subject_id is character type for merging
  mutate(clean_subject_id = as.character(clean_subject_id))

# Check for duplicate subject IDs after filtering, which could cause issues
if (any(duplicated(demographic_processed$clean_subject_id))) {
  warning("Duplicate 'clean_subject_id' found in demographic data after filtering. This may cause issues during merging.")
}
```

# Filter Atlas Data and Merge with Demographic Information
```{r}
# Get the exact column names expected from the template (excluding ID and AGE for now)
template_parcellation_cols <- setdiff(colnames(template_brainage), c("AGE", "clean_subject_id")) # Assuming template doesn't have clean_subject_id

# Filter atlas data for subjects present in your demographic subset
gendered_atlas_data <- atlas_processed %>%
  filter(clean_subject_id %in% demographic_processed$clean_subject_id) %>%
  # Select only the clean_subject_id and the relevant brain measure columns
  select(clean_subject_id, all_of(template_parcellation_cols[template_parcellation_cols %in% colnames(.)])) %>%
  # Ensure brain measures are numeric where possible before merging
  mutate(across(where(is.character) & !matches("clean_subject_id"), ~ suppressWarnings(as.numeric(.))))

# Start building the final data frame with subject IDs and their ages
final_template <- demographic_processed %>%
  select(clean_subject_id, AGE) %>%
  # Merge the brain measure data into final_template by clean_subject_id
  left_join(gendered_atlas_data, by = "clean_subject_id")

# Define the *exact* target column order from the original template
# The `template_brainage.xlsx` does NOT have `clean_subject_id`.
# The final output to the calculator should match the template exactly.
# Using `template_brainage` column names for final selection and order.
final_target_columns <- colnames(template_brainage)

# Add any columns from the template that are missing in final_template, filled with NA
# This ensures that even if a parcellation area has no data for the subjects, its column exists.
missing_in_final_template <- setdiff(final_target_columns, colnames(final_template))
if (length(missing_in_final_template) > 0) {
  final_template[missing_in_final_template] <- NA
}

# Reorder final_template columns to match the template_brainage order exactly
# Using `all_of` for robust selection of columns by name.
final_template <- final_template %>%
  select(all_of(final_target_columns))

# Verify that all target columns are present in the final data
columns_not_present_in_final <- setdiff(final_target_columns, colnames(final_template))
if(length(columns_not_present_in_final) == 0){
  message("All target columns from template_brainage are present in final_template.")
} else {
  warning(paste("The following target columns are still missing from final_template:", 
                paste(columns_not_present_in_final, collapse = ", ")))
}

# Display first few rows of the prepared data for review
print("--- Head of final_template (before final numeric conversion and ID removal for calculator) ---")
print(head(final_template))

print("--- Column names of final_template ---")
print(colnames(final_template))
```

# Output processed data
```{r}
# Define output directory and ensure path is expanded
output_dir <- path.expand("~/Developmental_BrainAGE/output/")

# Ensure output directory exists
if (!dir.exists(output_dir)) {
  dir.create(output_dir, recursive = TRUE)
}

# Output a .csv file that contains the subject ID and the age (for reference)
# This includes the `clean_subject_id` column.
output_csv_file <- paste0(output_dir, tolower(sex_filter), "_subID_developmental_template.csv")
write_csv(final_template, file = output_csv_file)
print(paste("Data with subject IDs saved as:", output_csv_file))

# Prepare data for the BrainAGE calculator: remove subject IDs and convert all to numeric.
# The template_brainage.xlsx does NOT have a subject ID column.
numeric_data_for_calculator <- final_template %>%
  # Select only columns that are *also* in the original template_brainage (excluding ID)
  select(all_of(colnames(template_brainage))) %>%
  # Convert everything to numeric. Suppress warnings for NA coercion (e.g., if a column was all NA)
  mutate(across(everything(), ~ suppressWarnings(as.numeric(as.character(.))))) # Convert to char first to avoid issues with some types

# Define output Excel file path
output_excel_file <- paste0(output_dir, tolower(sex_filter), "_populated_developmental_template.xlsx")

# Write the Excel file
write.xlsx(numeric_data_for_calculator, output_excel_file)

# Print confirmation with full path
print(paste("Template has been populated and saved as", output_excel_file))

print("Script finished successfully. Please check the output directory for your files.")
```


