---
title: "HBN_data"
output: html_document
date: "2025-05-21"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

This markdown includes steps to download the HBN dataset from reproducible braincharts (https://reprobrainchart.github.io/docs/get_data). These steps can be applied to any of the RBC datasets but they would need to be modified to include the respective dataset name. 

## Create directories, download the data using datalad, and then get the necessary files. Warning: do not try to get all the files as the dataset is very large and it would take up a lot of space on your hardrive. 

```{bash}
#Make a data directory to hold your data files
mkdir data
cd data

# Clone the HBN directory into the data directory
  datalad clone https://github.com/ReproBrainChart/HBN_FreeSurfer.git \
    -b complete-pass-0.1

#Download all .tsv files
cd HBN_FreeSurfer/freesurfer
find . -name "*.tsv" -exec datalad get {} +

#Make a directory to store all the .tsv files

mkdir -p ../HBN_tsv

#Make a sub-drectory for the brainmeasures tsv and the regional surface stats tv
cd HBN_tsv
mkdir brainmeasures_tsv
mkdir regionsurfacestats_tsv

find . -name "*regionsurfacestats.tsv" -exec cp {} HBN_tsv/regional_stats_tsv/ \;
find . -name "*brainmeasures.tsv" -exec cp {} HBN_tsv/brainmeasures_tsv/ \;
```


## Create group tsv files for the brainmeasures tsv with all the subject IDs combined into one file. This will make the data easier to use for the brainage calculators. 

```{r}
library(tidyverse)

# Set the directory containing the .tsv files
brainmeasures_tsv_dir <- "~/Data/HBN_tsv/brainmeasures_tsv"

# List all .tsv files
brainmeasures_tsv_files <- list.files(brainmeasures_tsv_dir, pattern = "\\.tsv$", full.names = TRUE)

# Read and combine all files without storing the filename
combined_df <- map_df(brainmeasures_tsv_files, ~ read_tsv(.x, show_col_types = FALSE))

# Preview combined data
print(combined_df)

# Create a new csv file with the combined data 
write_csv(combined_df, "~/Data/HBN_tsv/combined_brainmeasures.csv")

```
## Create a group tsv file for regional surface data

```{r}
# Set the directory containing the .tsv files
regionsurfacestats_tsv_dir <- "~/Data/HBN_tsv/regionsurfacestats_tsv"

# List all .tsv files
regionsurfacestats_tsv_files <- list.files(regionsurfacestats_tsv_dir, pattern = "\\.tsv$", full.names = TRUE)

# Read and combine all files without storing the filename
regions_combined_df <- map_df(regionsurfacestats_tsv_files, ~ read_tsv(.x, show_col_types = FALSE))

# Preview combined data
print(regions_combined_df)

# Create a new csv file with the combined data 
write_csv(regions_combined_df, "/Users/jiyashah/Data/HBN_tsv/combined_regionsurfacestats.csv")
```

## Get a list of atlases 

We want a list of atlas names because we want to create seperate tsv files for each atlas. This can be done by first creating a list and saving a list of unique atlases

```{r}
atlas_list <- unique(regions_combined_df$atlas)

print(atlas_list)

```
## Function to reshape the data

We then want to change the regional data into wide format and seperate each atlas so that the specific atlas files can be used when needed 

```{r}
library(tidyverse)

reshape_atlas_to_wide <- function(df, atlas_name) {
  df %>%
    # Keep only one atlas at a time
    filter(atlas == atlas_name, !is.na(StructName), StructName != "") %>%
    
    # Create new composite column name: hemisphere + region + metric
    pivot_longer(
      cols = -c(subject_id, session_id, atlas, hemisphere, StructName),
      names_to = "metric", values_to = "value"
    ) %>%
    mutate(variable = paste(hemisphere, StructName, metric, sep = "_")) %>%
    select(subject_id, session_id, variable, value) %>%
    
    # Aggregate (just in case) and reshape to wide
    group_by(subject_id, session_id, variable) %>%
    summarise(value = mean(value, na.rm = TRUE), .groups = "drop") %>%
    pivot_wider(names_from = variable, values_from = value)
}

```

## Apply the function

This will apply the function and output the combined csv files in the output directory 

```{r}
# Directory to save reshaped CSVs
output_dir <- "/Data/"
dir.create(output_dir, showWarnings = FALSE)

walk(atlas_list, function(atlas_name) {
  message("Reshaping: ", atlas_name)
  wide_df <- reshape_atlas_to_wide(regions_combined_df, atlas_name)
  write_csv(wide_df, paste0(atlas_name, "_wide.csv"))
})

```

