---
title: "Kauffman_brainage"
output: html_document
date: "2025-05-28"
---

## Load necessary libraries

```{r}
library(readxl)
library(openxlsx)
library(stringr)
library(dplyr)
library(readr)
library(xgb)
```

## R Markdown

This is a script to use the kauffman brainage calcualtor: https://github.com/tobias-kaufmann/brainage/tree/master

```{bash}
git clone https://github.com/tobias-kaufmann/brainage.git

```

## Load all the necessary files + change feature names file to the right format

```{r cars}
feature_names <- read_csv("brainage/feature-names.csv")
combined_brainmeasures <- read_csv("~/Data/HBN_tsv/combined_brainmeasures.csv")
glassr_output <- read_csv("~/Data/glasser_wide.csv")
demographic_df <- read_tsv("~/Data/HBN_BIDS/study-HBN_desc-participants.tsv")

```

<<<<<<< Updated upstream
<<<<<<< Updated upstream

=======
>>>>>>> Stashed changes
=======
>>>>>>> Stashed changes
```{r}
clean_subject_ids_tsv <- function(df) {
  colnames(df)[1] <- "participant_id"  # Rename the first column
  df <- df %>%
    mutate(clean_subject_id = str_remove(participant_id, "^sub-"))  # Remove "sub-" from start
  return(df)
}

brainmeasures_clean <- clean_subject_ids_tsv(combined_brainmeasures)
glassr_clean <- clean_subject_ids_tsv(glassr_output)
<<<<<<< Updated upstream
<<<<<<< Updated upstream


=======
>>>>>>> Stashed changes
=======
>>>>>>> Stashed changes
```


## Rename column names

```{r pressure, echo=FALSE}
colnames(glassr_clean) <- colnames(glassr_clean) %>%
  str_replace_all("ThickAvg", "thickness") %>%
  str_replace_all("SurfArea", "area") %>%
  str_replace_all("GrayVol", "volume") %>%
  str_replace_all("-", ".")

```


```{r}
<<<<<<< Updated upstream
<<<<<<< Updated upstream
# 1. Get the column name (which is also a feature)
col_header <- colnames(feature_names)[1]

# 2. Get the actual values
col_values <- feature_names[[1]]

# 3. Combine them
all_features <- c(col_header, col_values)

# 4. Add "clean_subject_id" at the front
column_names <- c("clean_subject_id", all_features)

# 5. Create empty dataframe
=======
=======
>>>>>>> Stashed changes
# Assume feature_names is a vector or single-column data frame
column_names <- feature_names[[1]]  # extract the vector of names

# Optionally, prepend a subject_id column
column_names <- c("clean_subject_id", column_names)

# Create an empty dataframe with those column names and 0 rows
<<<<<<< Updated upstream
>>>>>>> Stashed changes
=======
>>>>>>> Stashed changes
wide_structures <- data.frame(matrix(nrow = 0, ncol = length(column_names)))
colnames(wide_structures) <- column_names

# Check it
print(wide_structures)
```
<<<<<<< Updated upstream
<<<<<<< Updated upstream

=======
>>>>>>> Stashed changes
=======
>>>>>>> Stashed changes
```{r}
gendered_data <- demographic_df %>%
  filter(sex == "Female") %>%
  select(clean_subject_id = participant_id, sex)

print(gendered_data)
```

```{r}
gendered_glassr_data <- glassr_clean %>%
  filter(clean_subject_id %in% gendered_data$clean_subject_id)

print(gendered_glassr_data)
```

## Rename the brainmeasures to match the templates

```{r}
<<<<<<< Updated upstream
<<<<<<< Updated upstream
brainmeasures_renamed <- brainmeasures_clean %>%
  rename(
    Left.Lateral.Ventricle = Left_Lateral_Ventricle_Volume_mm3,
    Left.Inf.Lat.Vent = Left_Inf_Lat_Vent_Volume_mm3,
    Left.Cerebellum.White.Matter = Left_Cerebellum_White_Matter_Volume_mm3,
    Left.Cerebellum.Cortex = Left_Cerebellum_Cortex_Volume_mm3,
    Left.Thalamus.Proper = Left_Thalamus_Proper_Volume_mm3,
    Left.Caudate = Left_Caudate_Volume_mm3,
    Left.Putamen = Left_Putamen_Volume_mm3,
    Left.Pallidum = Left_Pallidum_Volume_mm3,
    X3rd.Ventricle = Third_Ventricle_Volume_mm3,
    X4th.Ventricle = Fourth_Ventricle_Volume_mm3,
    Brain.Stem = Brain_Stem_Volume_mm3,
    Left.Hippocampus = Left_Hippocampus_Volume_mm3,
    Left.Amygdala = Left_Amygdala_Volume_mm3,
    Left.Accumbens.area = Left_Accumbens_area_Volume_mm3,
    Right.Lateral.Ventricle = Right_Lateral_Ventricle_Volume_mm3,
    Right.Inf.Lat.Vent = Right_Inf_Lat_Vent_Volume_mm3,
    Right.Cerebellum.White.Matter = Right_Cerebellum_White_Matter_Volume_mm3,
    Right.Cerebellum.Cortex = Right_Cerebellum_Cortex_Volume_mm3,
    Right.Thalamus.Proper = Right_Thalamus_Proper_Volume_mm3,
    Right.Caudate = Right_Caudate_Volume_mm3,
    Right.Putamen = Right_Putamen_Volume_mm3,
    Right.Pallidum = Right_Pallidum_Volume_mm3,
    Right.Hippocampus = Right_Hippocampus_Volume_mm3,
    Right.Amygdala = Right_Amygdala_Volume_mm3,
    Right.Accumbens.area = Right_Accumbens_area_Volume_mm3,
    CC_Posterior = CC_Posterior_Volume_mm3,
    CC_Mid_Posterior = CC_Mid_Posterior_Volume_mm3,
    CC_Central = CC_Central_Volume_mm3,
    CC_Mid_Anterior = CC_Mid_Anterior_Volume_mm3,
    CC_Anterior = CC_Anterior_Volume_mm3,
    lhCortexVol = lhCortex_lhCortexVol,
    rhCortexVol = rhCortex_rhCortexVol,
    lhCorticalWhiteMatterVol = lhCerebralWhiteMatter_lhCerebralWhiteMatterVol,
    rhCorticalWhiteMatterVol = rhCerebralWhiteMatter_rhCerebralWhiteMatterVol,
    SubCortGrayVol = SubCortGray_SubCortGrayVol,
    TotalGrayVol = TotalGray_TotalGrayVol,
    SupraTentorialVol = SupraTentorial_SupraTentorialVol,
    EstimatedTotalIntraCranialVol = EstimatedTotalIntraCranialVol_eTIV
  )

=======
=======
>>>>>>> Stashed changes
# Define structure name mapping
structure_map <- c(
  "Left_Lateral_Ventricle" = "Left.Lateral.Ventricle",
  "Left_Inf_Lat_Vent" = "Left.Inf.Lat.Vent",
  "Left_Cerebellum_White_Matter" = "Left.Cerebellum.White.Matter",
  "Left_Cerebellum_Cortex" = "Left.Cerebellum.Cortex",
  "Left_Thalamus_Proper" = "Left.Thalamus.Proper",
  "Left_Caudate" = "Left.Caudate",
  "Left_Putamen" = "Left.Putamen",
  "Left_Pallidum" = "Left.Pallidum",
  "Third_Ventricle" = "X3rd.Ventricle",
  "Fourth_Ventricle" = "X4th.Ventricle",
  "Brain_Stem" = "Brain.Stem",
  "Left_Hippocampus" = "Left.Hippocampus",
  "Left_Amygdala" = "Left.Amygdala",
  "Left_Accumbens_area" = "Left.Accumbens.area",
  "Right_Lateral_Ventricle" = "Right.Lateral.Ventricle",
  "Right_Inf_Lat_Vent" = "Right.Inf.Lat.Vent",
  "Right_Cerebellum_White_Matter" = "Right.Cerebellum.White.Matter",
  "Right_Cerebellum_Cortex" = "Right.Cerebellum.Cortex",
  "Right_Thalamus_Proper" = "Right.Thalamus.Proper",
  "Right_Caudate" = "Right.Caudate",
  "Right_Putamen" = "Right.Putamen",
  "Right_Pallidum" = "Right.Pallidum",
  "Right_Hippocampus" = "Right.Hippocampus",
  "Right_Amygdala" = "Right.Amygdala",
  "Right_Accumbens_area" = "Right.Accumbens.area",
  "CC_Posterior" = "CC_Posterior",
  "CC_Mid_Posterior" = "CC_Mid_Posterior",
  "CC_Central" = "CC_Central",
  "CC_Mid_Anterior" = "CC_Mid_Anterior",
  "CC_Anterior" = "CC_Anterior",
  "lhCortex" = "lhCortexVol",
  "rhCortex" = "rhCortexVol",
  "lhCerebralWhiteMatter" = "lhCorticalWhiteMatterVol",
  "rhCerebralWhiteMatter" = "rhCorticalWhiteMatterVol",
  "SubCortGray" = "SubCortGrayVol",
  "TotalGray" = "TotalGrayVol",
  "SupraTentorial" = "SupraTentorialVol",
  "EstimatedTotalIntraCranialVol" = "EstimatedTotalIntraCranialVol"
)

# Function to apply mapping and ensure unique column names
rename_columns_no_duplicates <- function(cols, map) {
  new_names <- character(length(cols))
  name_tracker <- list()
  
  for (i in seq_along(cols)) {
    col <- cols[i]
    renamed <- col  # default
    matched <- FALSE

    for (prefix in names(map)) {
      if (startsWith(col, prefix)) {
        matched <- TRUE
        base_name <- map[[prefix]]

        if (grepl("_Volume_mm3$", col)) {
          # Keep one clean name for _Volume_mm3
          if (is.null(name_tracker[[base_name]])) {
            renamed <- base_name
            name_tracker[[base_name]] <- 1
          } else {
            name_tracker[[base_name]] <- name_tracker[[base_name]] + 1
            renamed <- paste0(base_name, ".", name_tracker[[base_name]])
          }
        } else {
          # All others get suffixes
          if (is.null(name_tracker[[base_name]])) {
            name_tracker[[base_name]] <- 2
          } else {
            name_tracker[[base_name]] <- name_tracker[[base_name]] + 1
          }
          renamed <- paste0(base_name, ".", name_tracker[[base_name]])
        }
        break
      }
    }

    new_names[i] <- if (matched) renamed else col
  }

  # Ensure uniqueness
  make.unique(new_names, sep = ".")
}

# Apply renaming to your dataframe
brainmeasures_renamed <- brainmeasures_clean
colnames(brainmeasures_renamed) <- rename_columns_no_duplicates(colnames(brainmeasures_clean), structure_map)
<<<<<<< Updated upstream
>>>>>>> Stashed changes
=======
>>>>>>> Stashed changes

```

## Only get the female data from the brain measures file

```{r}
gendered_brainmeasure_data <- brainmeasures_renamed%>%
  filter(clean_subject_id %in% gendered_data$clean_subject_id)

print(gendered_brainmeasure_data)
```

```{r}
gendered_data$clean_subject_id <- as.character(gendered_data$clean_subject_id)
gendered_brainmeasure_data$clean_subject_id <- as.character(gendered_brainmeasure_data$clean_subject_id)

# Save the original column names (structure) before doing anything
# First, check if template_df has rows
if (nrow(wide_structures) == 0) {
  # Create an empty data frame with the same structure as template_df but add the subject IDs and age
<<<<<<< Updated upstream
<<<<<<< Updated upstream
=======
  # Keep all the original columns, but add the new data
  wide_structures <- wide_structures[0, ]  # Retain column structure but no rows
>>>>>>> Stashed changes
=======
  # Keep all the original columns, but add the new data
  wide_structures <- wide_structures[0, ]  # Retain column structure but no rows
>>>>>>> Stashed changes
  
  # Now create a new data frame that only updates clean_subject_id and AGE
  new_data <- data.frame(clean_subject_id =gendered_data$clean_subject_id,
                         stringsAsFactors = FALSE)
  
<<<<<<< Updated upstream
<<<<<<< Updated upstream
  new_data$clean_subject_id <- as.character(new_data$clean_subject_id)
  wide_structures$clean_subject_id <- as.character(wide_structures$clean_subject_id)
  
=======
>>>>>>> Stashed changes
=======
>>>>>>> Stashed changes
  # Bind the new data to the empty template, ensuring columns remain intact
  wide_structures <- bind_rows(wide_structures, new_data)
} else {
  # If template_df already has rows, just update clean_subject_id and AGE columns
  wide_structures$clean_subject_id <- gendered_data$clean_subject_id
}

print(wide_structures)
```


<<<<<<< Updated upstream
<<<<<<< Updated upstream
```{r}
# Convert all columns in the template_df to characters
wide_structures <- wide_structures %>%
  mutate(across(everything(), as.character))

gendered_brainmeasure_data <- gendered_brainmeasure_data %>%
  mutate(across(everything(), as.character))

gendered_glassr_data <- gendered_glassr_data %>%
  mutate(across(everything(), as.character))

```

```{r}
final_template <- left_join(wide_structures[, c(1), drop = FALSE], gendered_glassr_data, by = "clean_subject_id") %>%  
  left_join(gendered_brainmeasure_data, by = "clean_subject_id") 

# Select only the columns from template_df
final_template <- final_template %>%
  select(names(wide_structures))  # Keeps only the columns in templa5te_df  

# Remove columns with NA
final_template <- final_template %>%
  filter(!is.na(final_template[[3]]))

print(final_template)
```

```{r}
#Check to see if the two column names match

# Compare column names between two dataframes
common_cols <- intersect(names(final_template), names(wide_structures))

# See how many columns match
length(common_cols)

# (Optional) Print which columns match
print(common_cols)
```
```{r}
#remove subject Id column
features_females <- final_template[, !(names(final_template) %in% "clean_subject_id")]

print(features_females)
```

=======
=======
>>>>>>> Stashed changes
# Check if subject IDs from both dataframes match

```{r}
common_ids <- intersect(template_df$clean_subject_id, gendered_lh_thickness_data$clean_subject_id)
print(paste("Number of matching subject IDs:", length(common_ids)))
<<<<<<< Updated upstream
```
>>>>>>> Stashed changes
=======
```
>>>>>>> Stashed changes
