---
title: "NAPR_analsysis"
output: html_document
---

# Overview
This R Markdown script creates and populates NAPR BrainAGE templates using freesurfer outputs.

Files and column names have been adapted to align with RBC conventions.

## Background: TIGRLab NAPR Framework

NAPR (Neuroimaging Age Prediction using Regression) is a cloud-based framework developed by TIGRLab for estimating brain age from structural MRI data. It uses Freesurfer-processed surfaces mapped to the fsaverage4 template as input and applies predictive models like Gaussian Process Regression (gausspr) and Relevance Vector Machine (rvmm).

The original software ran on www.cloudneuro.org (now discontinued), and the codebase is available on GitHub: [https://github.com/TIGRLab/napr](https://github.com/TIGRLab/napr).

This script processes the NAPR output files alongside demographic and quality control data to facilitate brain age analyses following RBC standards.


```{r setup, include=FALSE}

# Load packages
library(ggplot2)
library(readr)
library(readxl)
library(rstatix)
library(broom)
library(stringr)
library(dplyr)
```

```{bash}
#Create directory to store files for BrainAGE calculation

# Create the main project directory NAPR_BrainAGE folder
mkdir -p NAPR_BrainAGE
cd /path/to/your/NAPR_BrainAGE

#make folders within NAPR_BrainAGE folder
mkdir -p NAPR_BrainAGE/Freesurfer_Outputs
mkdir -p NAPR_BrainAGE/output

# Confirm the structure
tree NAPR_BrainAGE

#Download the NAPR BrainAGE template, project demographic data,freesurfer outputs, datacut IDS and euler data

# Move your input files into the working directory (update file names as needed)
mv ~/Downloads/<NAPR_template_file.xlsx> NAPR_BrainAGE/
mv ~/Downloads/<demographics_file.tsv> NAPR_BrainAGE/
mv ~/Downloads/<datacut_ids_file.csv> NAPR_BrainAGE/
mv ~/Downloads/<euler_file.tsv> NAPR_BrainAGE/


# Move freesurfer files into subfolder
mv ~/Downloads/lh.Schaefer2018_1000Parcels.*.tsv NAPR_BrainAGE/Freesurfer_Outputs/
mv ~/Downloads/rh.Schaefer2018_1000Parcels.*.tsv NAPR_BrainAGE/Freesurfer_Outputs/
```

#Load and Combine NAPR Output files

```{r pressure, echo=FALSE}
# Define path to directory containing NAPR output .txt files
library(dplyr)
file_dir <- "NAPR_BrainAGE/output/" 

# List all text files in the directory
napr_files <- list.files(file_dir, pattern = "\\.txt$", full.names = TRUE)

# Read and combine all NAPR output files into a single dataframe
combined_data <- do.call(rbind, lapply(napr_files, function(file) {

# Read the file
data <- read.table(file, header = TRUE, sep = "\t", stringsAsFactors = FALSE)
  
# Extract the subject name from the file name
subject_name <- gsub(".txt", "", basename(file))  # Remove ".txt" from file name
  
# Add the subject name as a new column
data$clean_subject_id <- subject_name
  
  return(data)  # Return the modified data
}))

print(combined_data)

```

#Process combined NAPR data

```{r}
library(tidyr)
# Assuming combined_data is your dataframe

NAPR_combined_data <- combined_data %>%
  separate(Model..........................Predicted.age, into = c("Model", "Predicted.age"), sep = "\\s+", convert = TRUE) %>%

# Pivot wider to create separate columns for gausspr and rvmm
  pivot_wider(
    id_cols =  clean_subject_id,            # Keep subject_name as the unique identifier
    names_from = Model,                # Use Model as column names
    values_from = Predicted.age        # Use Predicted.age as values
  ) %>%
  
  # Rename columns for clarity
  rename(
    gausspr_age = `gausspr.full.20161228`,
    rvmm_age = `rvmm.full.20161228`
  )

print(NAPR_combined_data)
```

#Demographic Data

```{r}
#Load demographic data

demographics <- read_csv("NAPR_BrainAGE/demographics.tsv", show_col_types = FALSE)

# Load datacut IDs
datacut_ids <- read_csv("NAPR_BrainAGE/datacut_ids.csv", show_col_types = FALSE)

# Load Euler number QC metric

euler_data <- read_tsv("NAPR_BrainAGE/euler.tsv")
```

#Clean subject IDs in demographic and QC data

```{r}
clean_subject_ids_demo <- function(df, subject_id) {
  df %>%
    mutate(clean_subject_id = str_extract(!!sym(subject_id), "[0-9]{8}$"))
}

demographics_clean <- clean_subject_ids_demo(demographics, "subject_id")

datacut_ids_clean <- clean_subject_ids_demo(datacut_ids, "subject_id")
```

```{r}
clean_subject_ids_tsv <- function(df) {
  colnames(df)[1] <- "clean_subject_id"  # Rename the first column to 'clean_subject_id'
  df %>%
    mutate(clean_subject_id = str_extract(clean_subject_id, "[0-9]{8}$"))  # Extract the last 8 digits of subject IDs
}

euler_data_clean <- clean_subject_ids_tsv(euler_data)

print(euler_data_clean)
```

#Clean the NAPR combined file subject IDS

```{r}
# Assuming your dataframe is named 'combined_data' and the column is 'subject_name'

NAPR_combined_data <- NAPR_combined_data %>%
  mutate(clean_subject_id = gsub("\\D", "", clean_subject_id))  # Replace non-digit characters with ""

# View the updated dataframe
print(NAPR_combined_data)

```

#Convert subject IDs to numeric for merging

```{r}
demographics_clean$clean_subject_id <- as.numeric(demographics_clean$clean_subject_id)
datacut_clean$clean_subject_id <- as.numeric(datacut_clean$clean_subject_id)
euler_clean$clean_subject_id <- as.numeric(euler_clean$clean_subject_id)
NAPR_combined_data$clean_subject_id <- as.numeric(NAPR_combined_data$clean_subject_id)
```

# Prioritize demographic rows where symptoms are filled out
```{r}
# Group by subject ID and prioritize rows where symptoms are filled out
# If symptoms are NA, the function will pick the first occurrence that has them filled
unique_symptoms <- demographics_clean %>%
  group_by(clean_subject_id) %>%
  arrange(desc(!is.na(prodromal_psychosis))) %>%  # Arrange to prioritize rows where symptoms are NOT NA
  slice(1) %>%  # Keep the first row per subject 
  ungroup()

# The result will be a dataframe with unique subject IDs and prioritized symptom data
print(unique_symptoms)
```

#Merge data for final analysis

```{r}
# Select relevant columns from NAPR predictions
final_NAPR_analysis_data <- NAPR_combined_data %>%
  select(clean_subject_id, rvmm_age, gausspr_age)

# Join with unique symptom data (prodromal psychosis, age at scan, sex assigned at birth)
final_NAPR_analysis_data <- final_NAPR_analysis_data %>%
  left_join(unique_symptoms %>%
              select(clean_subject_id, prodromal_psychosis, age_scan, assigned_sex_at_birth),
            by = "clean_subject_id")
print(final_NAPR_analysis_data)
```

```{r}
# Add Euler number (quality metric) to the dataset
final_NAPR_analysis_data <- final_NAPR_analysis_data %>%
  left_join(euler_data_clean %>% 
              select(clean_subject_id, mean_euler_bh),  
            by = "clean_subject_id")
```

```{r}
# Assuming 'sub_ID' is the column to match in both data frames and that `t01` is a logical or character column.
final_NAPR_analysis_data <- final_NAPR_analysis_data %>%
  # Join with datacut_ids_clean to bring in the 't01' column
  left_join(datacut_ids_clean %>% select(clean_subject_id, t01), by = "clean_subject_id")

# View the result
print(final_NAPR_analysis_data)

datacut_filtered_data <- final_NAPR_analysis_data %>%
  filter(t01 == TRUE)

# View the result
print(datacut_filtered_data)
```

```{r}
# Further filter for good surface reconstruction quality (Euler number threshold)
euler_filtered_data <- datacut_filtered_data %>%
  filter(mean_euler_bh >= -200)

print(euler_filtered_data)
```

```{r}
# Optional: Apply manual corrections for known data issues
# euler_filtered_data <- euler_filtered_data %>%
#   mutate(sex = case_when(
#     participant_id == "12345678" ~ "Male",
#     TRUE ~ sex
#   ))

# Identify any remaining rows with missing values
rows_with_na <- euler_filtered_data %>%
  filter(if_any(everything(), is.na))
print(rows_with_na)
```

```{r}
cleaned_filtered_data <- na.omit(euler_filtered_data)

cleaned_filtered_data <- cleaned_filtered_data[cleaned_filtered_data$clean_subject_id != "ID Number", ]

# Rename columns to match RBC conventions
cleaned_filtered_data <- cleaned_filtered_data %>%
  rename(
    participant_id = clean_subject_id,
    age = age_scan,
    sex = assigned_sex_at_birth
  )
print(cleaned_filtered_data)

write.csv(cleaned_filtered_data, "NAPR_BrainAGE/output/complete_NAPR_data.csv", row.names = FALSE)
```

## References

- Franke, K., & Gaser, C. (2012). *Longitudinal changes in individual BrainAGE...* GeroPsych, 25(4), 235–245. [https://doi.org/10.1024/1662-9647/a000074](https://doi.org/10.1024/1662-9647/a000074)
- Pardoe, H. R. et al. (2017). *Structural brain changes in epilepsy resemble premature brain aging*. Epilepsy Research, 133, 28–32. [https://doi.org/10.1016/j.eplepsyres.2017.04.001](https://doi.org/10.1016/j.eplepsyres.2017.04.001)
- NAPR GitHub: [https://github.com/hpardoe/napr](https://github.com/hpardoe/napr)
- Bethlehem, R. A. I., et al. (2022). *Brain charts for the human lifespan*. Nature. [https://doi.org/10.1038/s41586-022-04554-y](https://doi.org/10.1038/s41586-022-04554-y)

- Cole JH, Franke K. Predicting Age Using Neuroimaging: Innovative Brain Ageing Biomarkers. Trends Neurosci. 2017;40(12):681–690.  
- Pardoe HR et al., NAPR: A cloud-based framework for age prediction using neuroimaging. Neuroimage. 2016.

