---
title: "Network_brain_age"
output: html_notebook
date: 2024-10-18
---

# Overview

This is a script to create and populate the Network BrainAge template (https://centilebrain.org/#/brainAGE) using freesurfer outputs.

Files and column names have been adapted to align with RBC conventions.

```{r}
## Load Required Libraries
library(readxl)
library(openxlsx)
library(stringr)
library(dplyr)
library(readr)
library(curl)
```

```{bash}
#Create directory to store files for BrainAGE calculation

# Create the main project directory Network_BrainAGE folder
mkdir -p Network_BrainAGE
cd /path/to/your/Network_BrainAGE

#make folders within Network_BrainAGE folder
mkdir -p Network_BrainAGE/Freesurfer_Outputs
mkdir -p Network_BrainAGE/output

# Confirm the structure
tree Network_BrainAGE

#Download the Network BrainAGE template, project demographic data, and freesurfer outputs

# Move template and demographic files into main project folder
mv ~/Downloads/Network_BrainAGE_template.xlsx Network_BrainAGE/
mv ~/Downloads/demographics.tsv Network_BrainAGE/

# Move freesurfer files into subfolder
mv ~/Downloads/lh.Schaefer2018_1000Parcels.*.tsv Network_BrainAGE/Freesurfer_Outputs/
mv ~/Downloads/rh.Schaefer2018_1000Parcels.*.tsv Network_BrainAGE/Freesurfer_Outputs/
```

Load Files
# Change sex to "Male" or "Female" depending on analysis.

```{r}
# Set working directory to main project folder before running this script
setwd("Network_BrainAGE")

#Load Network BrainAGE template and demographic data
template_brainage <- read_excel("Network_BrainAGE_template.xlsx")
demographics_df <- read_tsv("demographics.tsv")

# Filter for Sex ('Male' or 'Female')
sex <- "Male"

# Load Schaefer 1000-parcel thickness and surface area data

lh_thick <- read_tsv("Freesurfer_Outputs/lh.Schaefer2018_1000Parcels.thickness.tsv")
lh_surf  <- read_tsv("Freesurfer_Outputs/lh.Schaefer2018_1000Parcels.surfacearea.tsv")
rh_thick <- read_tsv("Freesurfer_Outputs/rh.Schaefer2018_1000Parcels.thickness.tsv")
rh_surf  <- read_tsv("Freesurfer_Outputs/rh.Schaefer2018_1000Parcels.surfacearea.tsv")
```

# Clean and Standardize subject IDs

```{r}
# Subject IDs in cortical metrics data
clean_subject_ids_tsv <- function(df) {
  colnames(df)[1] <- "clean_subject_id"  # Rename the first column to 'clean_subject_id'
  df %>%
    mutate(clean_subject_id = str_extract(clean_subject_id, "[0-9]{8}$"))  # Extract the last 8 digits of subject IDs
}

#Function for demographics subject IDs
clean_subject_ids_demographics <- function(df, id_column) {
  df %>%
    mutate(clean_subject_id = str_extract(!!sym(id_column), "[0-9]{8}$"))
}
```

# Apply the subject_ID cleaning function to thickness and surface area files

```{r}
#Cortical metrics data
lh_thick_clean <- clean_subject_ids_tsv(lh_thick)
lh_surf_clean <- clean_subject_ids_tsv(lh_surf)
rh_thick_clean <- clean_subject_ids_tsv(rh_thick)
rh_surf_clean <- clean_subject_ids_tsv(rh_surf)

#demographics data
demographics_df <- clean_subject_ids_demographics(demographics_df, "participant_id")  # Adjust column name as per data
```

# Standardize the column name by renaming template subject ID column

```{r}
#Rename templates subject ID column for consistency
template_df <- template_brainage %>%
  rename(clean_subject_id = SubjectID)
```

# 1. Filters for time point 
# This also prevents duplicate subject_IDs in the final output

# 2. Filters for sex and then creates a dataframe with subject ID, age, and sex. 

```{r}

gendered_data <- demographics_df %>%
  filter(wave == "1") %>%  # Filter for the desired wave (baseline)
  filter(sex == sex) %>%  # Filter for sex
  select(clean_subject_id, age, sex)  # Keep subject ID, age, and sex for merging

```

# Filter cortical data for female subject IDs 

```{r}

filter_subjects <- function(metric_df, subject_ids) {
  metric_df %>%
    filter(clean_subject_id %in% subject_ids)
}

gendered_lh_thickness_data <- lh_thick_clean %>%
  filter(clean_subject_id %in% gendered_data$clean_subject_id)

gendered_lh_area_data <- lh_surf_clean %>%
  filter(clean_subject_id %in% gendered_data$clean_subject_id)

gendered_rh_thickness_data <- rh_thick_clean %>%
  filter(clean_subject_id %in% gendered_data$clean_subject_id)

gendered_rh_area_data <- rh_surf_clean %>%
  filter(clean_subject_id %in% gendered_data$clean_subject_id)
```

# Add the female subject_ID, age, and sex to the template_df
## This also ensures that there are enough rows to merge the data 

```{r}

# Make sure the subject_ids, sex, and age are character data types to ensure merge compatibility
template_df$clean_subject_id <- as.character(template_df$clean_subject_id)
template_df$sex <- as.character(template_df$sex)  # Add sex column conversion
template_df$age <- as.character(template_df$age)   # Add age column conversion

# Check if template_df has rows
if (nrow(template_df) == 0) {
  # Create an empty data frame with the same structure as template_df but add the subject IDs and age
  # Keep all the original columns, but add the new data
  template_df <- template_df[0, ]  # Retain column structure but no rows
  
  # Now create a new data frame that only updates clean_subject_id and AGE
  new_data <- data.frame(clean_subject_id =gendered_data$clean_subject_id,
                         age = gendered_data$age,
                         sex = gendered_data$sex,
                         stringsAsFactors = FALSE)
  
  # Make them the same data type
  new_data$clean_subject_id <- as.character(new_data$clean_subject_id)
  new_data$sex <- as.character(new_data$sex)
  new_data$age <- as.character(new_data$age)
  
  # Bind the new data to the empty template, ensuring columns remain intact
  template_df <- bind_rows(template_df, new_data)
} else {
  
  # If template_df already has rows, just update clean_subject_id and AGE columns
  template_df$clean_subject_id <- gendered_data$clean_subject_id
  template_df$AGE <- gendered_data$age
}

```

# Add the default values for the SITE (1 because they were all taken at the same site on the same scanner), ScannerType, and FreeSurfer_Version. Change values if needed. 

```{r}

# Create a named list of default values for columns with missing data
default_values <- list(
  SITE = "1",
  ScannerType = "3",
  FreeSurfer_Version = "7.0"
)

# Update existing columns with default values if they are NA
for (col in names(default_values)) {
  if (col %in% names(template_df)) {
    template_df[[col]] <- ifelse(is.na(template_df[[col]]), default_values[[col]], template_df[[col]])
  }
}

```


#Confirm Subject ID overlap

```{r}
# Check if subject IDs from both dataframes match
common_ids <- intersect(template_df$clean_subject_id, gendered_lh_thickness_data$clean_subject_id)
print(paste("Number of matching subject IDs:", length(common_ids)))
```

Convert logical columns to character

```{r}

# Convert all columns in the template_df to characters
template_df <- template_df %>%
  mutate(across(everything(), as.character))

gendered_data <- gendered_data %>%
  mutate(across(everything(), as.character))

gendered_lh_thickness_data <- gendered_lh_thickness_data %>%
  mutate(across(everything(), as.character))

gendered_lh_area_data <- gendered_lh_area_data %>%
  mutate(across(everything(), as.character))

gendered_rh_thickness_data <- gendered_rh_thickness_data %>%
  mutate(across(everything(), as.character))

gendered_rh_area_data <- gendered_rh_area_data %>%
  mutate(across(everything(), as.character))
```

# Merge Data into Final Template 
1. Populate the template with female subject data and only keep the column names that are in the template.
2. Remove any rows with NA, the calculator will not work if there are any NA values. 
3. Then revert back to the original column names. 
4. Ensure that all the relevant data is properly merged.

```{r}
final_template <- left_join(template_df[, c(1, 2, 3, 4, 5, 6)], gendered_lh_thickness_data, by = "clean_subject_id") %>%  
  left_join(gendered_lh_area_data, by = "clean_subject_id") %>%
  left_join(gendered_rh_thickness_data, by = "clean_subject_id") %>%
  left_join(gendered_rh_area_data, by = "clean_subject_id")

# Select only the columns from template_df
final_template <- final_template %>%
  select(names(template_df))  # Keeps only the columns in template_df  

#remove columns with NA
final_template <- final_template %>%
  filter(!is.na(final_template[[7]]))

```

#Convert Data to Numeric and remove first subject ID column (not in the original template)
## Make sure all the data types are numeric because the calculator will not work if they are different data types. 

```{r}

# Convert all other rows and columns to numeric
numeric_data <- final_template %>%
  mutate_at(vars(-4), ~ as.numeric(as.character(.)))

```

The Network brainAGE only accepts 150 participants at a time so multiple files need to be created to only have 150 participants per file.
# Output files in parts and then put them through the brain age calculator. 

```{r}
##Exporting data in chunkcs of 150 participants

participants_per_file <- 150

# Calculate how many files needed
num_files <- ceiling(nrow(numeric_data) / participants_per_file)

# Loop through the dataframe, slice into 150 participant chunks, and save each chunk
for (i in 1:num_files) {
  # Slice rows for participants
  start_row <- (i-1)*participants_per_file + 1
  end_row <- min(i*participants_per_file, nrow(numeric_data)) 
  
  chunked_participants <- numeric_data %>%
    slice(start_row:end_row)
  
  # Create file name for each chunk
  file_name <- paste0("output/", tolower(sex), "_network_template_part_", i, ".xlsx")
  
  # Write the chunk to an Excel file, keeping the header row
  write.xlsx(chunked_participants, file_name)
  
  # Print a message for each file creation
  print(paste("Saved file:", file_name))
}


```

Relevant references:
- Franke, K., & Gaser, C. (2012). Longitudinal changes in individual BrainAGE in healthy aging, mild cognitive impairment, and Alzheimer's disease. *GeroPsych: The Journal of Gerontopsychology and Geriatric Psychiatry*, 25(4), 235–245. [https://doi.org/10.1024/1662-9647/a000074](https://doi.org/10.1024/1662-9647/a000074)
- Bethlehem, R. A. I., et al. (2022). Brain charts for the human lifespan. *Nature*, 604(7906), 525–533. [https://doi.org/10.1038/s41586-022-04554-y](https://doi.org/10.1038/s41586-022-04554-y)

```{r}